
---
title: "P1.2 Clean Syphilis FHIR Data"
output:
  html_document:
    toc: true
---

Phase 1 - Notebook 2 (copied from Databricks, not all code may work correctly in other IDEs).

This notebook cleans and transforms newly downloaded FHIR data.

**Author:** Nate Bean

**Date created:** 11/23 -- **Last modified:** 11/23

**Purpose:** Transform syphilis data to match the original format after it's been converted from FHIR JSON. For now, do the remaining transformations in a third file.

**Schedule:** Daily


```{r}
remotes::install_github("ReverieNB/PHinfo")
```


```{r}
library(tidyverse)
library(PHinfo)
```


```{r}
raw <- "/dbfs/mnt/phmdw/Raw/PublicHealth/InfectiousDiseases/SyphilisProvisional/Daily Files/"
refined <- "/dbfs/mnt/phmdw/Refined/PublicHealth/InfectiousDiseases/SyphilisProvisional/"
geo <- "/dbfs/mnt/phmdw/Raw/PublicHealth/InfectiousDiseases/SyphilisProvisional/Geocoding/"
phcs <- "/dbfs/mnt/phmdw/Raw/PublicHealth/PHCS/Syphilis Provisional/"

new_file <- list.files(raw, pattern = "fhir_syphilis_data") |>
  tibble(file = _) |>
  mutate(date = ymd(str_remove_all(file, "^.*_|\\.csv"))) |>
  filter(date == max(date)) |>
  pull(file)

new_file
```


```{r}
raw_data <- read_csv(str_c(raw, new_file),
         #Select and rename columns
         col_select = c(
          case_id = id, case_modification_date, first_name, last_name, syphcode, age, age_units, birth_date, city, county_name = county, zip, disease, ethnc_code = ethnicity, 
          facility_doing_the_reporting, sex = gender, gender_detail, partner_female = Femalesexpartners, partner_male = Malesexpartners, 
          partner_transgender_fm = TransgenderFMsexpartners, partner_transgender_mf = TransgenderMFsexpartners, partner_unknown = `SexPartners'genderunknown`,
          preg = Pregnantattimeofdiagnosis, race, regimen_start_date1 = regimen_start_date, site, specimen_date, street_1, street_2, treatment_regimen),
         #Specify column types
         col_types = c(
          id = "c", 
          case_modification_date = "T",
          first_name = "c", 
          last_name = "c",
          syphcode = "c",
          age = "d",
          age_units = "c",
          birth_date = "D",
          city = "c",
          county = "c",
          zip = "c",
          disease = "c",
          ethnicity = "c",
          facility_doing_the_reporting = "c",
          gender = "c",
          gender_detail = "c",
          Femalesexpartners = "c",
          Malesexpartners = "c",
          TransgenderFMsexpartners = "c",
          TransgenderMFsexpartners = "c",
          "SexPartners'genderunknown" = "c",
          Pregnantattimeofdiagnosis = "c",
          race = "c",
          regimen_start_date = "D",
          site = "c",
          specimen_date = "D",
          street_1 = "c",
          street_2 = "c",
          treatment_regimen = "c"))
```


```{r}
#syphilis ytd has actual NA, while the previous files have placeholders or blank. what is preferred? going to match the 2022 format for now
aligned_data <- raw_data |>
  #Multiple values are concatenated in the query script -- separate
  separate_wider_delim(race, delim = ";;", names = c("race_code1", "race_code2", "race_code3"), too_few = "align_start") |> 
  separate_wider_delim(treatment_regimen, delim = ";;", names = c("treatment_regimen1", "treatment_regimen_other1"), too_few = "align_start") |> 
  separate_wider_delim(site, delim = ";;", names = c("site1", "site2"), too_few = "align_start") |> 
  #Convert variables to match their original formats/formats created in the original transformation script
  mutate(ethnc_code = case_when( 
          ethnc_code == "Hispanic or Latino" ~ 1,
          ethnc_code == "Non Hispanic or Latino" ~ 2,
          ethnc_code == "asked but unknown" ~ 9,
          ethnc_code == "Unknown" ~ 9),

        sex = case_when(
          sex == "male" ~ 1,
          sex == "female" ~ 2,
          sex == "other" ~ 9),

        gender_detail = case_when(
          gender_detail == "M_F" ~ "M-F",
          gender_detail == "F_M" ~ "F-M"),

        preg = case_when(
          preg == "YES" ~ 1,
          preg == "NO" ~ 2,
          preg == "UNK" ~ 9),

        syphcode = case_when(
          syphcode == "SYPH_PRIMARY" ~ 1,
          syphcode == "SYPH_SECONDARY" ~ 2,
          syphcode == "SYPH_EARLY_LATENT" ~ 3,
          syphcode == "SYPH_CONG" ~ 6,
          syphcode == "SYPH_UNKNOWN_OR_LATE" ~ 4), #map these cases to late_latent (4) original value

        across(starts_with("race_code"), ~case_when(            
          . == "White" ~ 1,
          . == "Black or African American" ~ 2,
          . == "American Indian or Alaska Native" ~ 3,
          . == "Asian" ~ 4,
          . == "Native Hawaiian or Other Pacific Islander" ~ 4, #combine into Asian/PI to match original
          . == "Other Race" ~ 5,
          . == "" | is.na(.) ~ 9)),

        across(starts_with("site"), ~case_when( #likely want to add a missing placeholder (9/99?) so NA returns an error here too
          as.numeric(.) == 6 ~ 10, #blood
          as.numeric(.) == 86 ~ 11, #CSV
          as.numeric(.) == 94 ~ 5,
          is.na(.) ~ NA)), #other

        across(starts_with("partner_"), ~if_else(. == "Y", 1, 9)),
        across(starts_with("street"), ~str_squish(.)),
        across(contains("treatment_reg"), ~str_to_title(.)),
        case_modification_date = as.Date(case_modification_date),
        zip = str_sub(zip, 1, 5),
        age_units = str_to_title(age_units))
```


```{r}
validated_data <- aligned_data |>
  #There are some cases where the same person (first, last, dob) has multiple rows/ids. In all cases, one record is more updated than the other
  #Idk why new records are created versus the old ones being updated - will reach out to MDH.
  #For now, just take the more recent one.
  mutate(identity = str_c(first_name, last_name, birth_date)) |>
  filter(case_modification_date == max(case_modification_date), .by = identity) |>
  filter(case_id == max(case_id), .by = identity) |>
  expect_values(list(sex = c(1, 2, 9),
                     race_code1 = c(1, 2, 3, 4, 5, 9),
                     race_code2 = c(1, 2, 3, 4, 5, 9),
                     race_code3 = c(1, 2, 3, 4, 5, 9),
                     ethnc_code = c(1, 2, 9),
                     syphcode = c(1, 2, 3, 4, 6),
                     county_name = "Hennepin County",
                     disease = "SYPHILIS")) |>
  expect_values(list(site1 = c(10, 11, 5)), na_accept = TRUE) |>
  expect_missing(list(case_id = 0, first_name = 0, last_name = 0, birth_date = 0, ethnc_code = 0, race_code1 = 0, race_code2 = 0, race_code3 = 0)) |>
  expect_unique("case_id") |>
  check_expectations(fail = TRUE) |>
  select(-identity)
```


```{r}
mini_skim(validated_data)
```


```{r}
#Before the FHIR process this was performed in a separate script
for_geo <- validated_data |>
  select(case_id, street_1, city, zip) |>
  fix_city("city")

#Print city names that were fixed
print(validated_data$city[!str_to_upper(validated_data$city) %in% for_geo$city])
```


```{r}
#Write out this list for Emily. She will use it to query Epic data that we will send to MDH as part of the Phase II project.
epic_list <- validated_data |>
  select(last_name, first_name, case_id, birth_date, specimen_date)

write_csv(epic_list, str_c(phcs, "syphilis_patients_for_match.csv"))
```


```{r}
#The data can't go through the Trusted data cleaning before being geocoded.
#Kelzee will do this manually, so that final stage of data cleaning is still manual.
write_rds(validated_data, str_c(refined, "fhir_syphilis_aligned.rds"))

#Also save data for geocoding - include all years in a single file
write_csv(for_geo, str_c(geo, "syphilis_ytd_for_geo.csv"))
```

